---
title: "GST Tutorial"
author:
  - name: Jiaxin Huang
output: rmarkdown::html_vignette
bibliography: references.bib  
vignette: >
  %\VignetteIndexEntry{GST Tutorial}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(GST)
library(knitr)
library(pwr)
```

## Introduction to the use of Global Statistical Approach

Determination whether one treatment is preferred over others is a major
goal of many clinical studies but can be complicated by the situation
when no single outcome is sufficient to make the judgement. This package
introduces a useful global statistics test technique and the
corresponding global treatment effect (GTE) measure, which employs a
nonparametric method based on a composite rank-sum of diverse variables
for assessing treatment's global preference when multiple outcomes are
evaluated together.

Measure of treatment benefit for each outcome is defined as
$\theta = p - q$ which is the difference between the probability that
the treatment is better than the control (p) and the probability that
the control is better than the treatment (q). When there are multiple
outcomes, compute such $\theta$ for each outcome and take an average
$\bar{\theta}$ of these values.This $\bar{\theta}$ value is called the
global treatment effect (GTE).

Interpretation of GTE is very straightforward: GTE = 0. indicates no
global difference between treatment 1 and 2, GTE \> 0 indicates
treatment 2 gives globally higher values than treatment 1 and GTE \< 0
indicates treamtent 2 gives globally smaller values than treatment 1. In
continuous case, GTE values approaching 1 implies an increased treatment
2's global values over treatment 1, whereas GTE values approaching -1
suggest a decreased treatment 2's global values over treatment 1.

This package provides a simple useful method to compare a treatment's
global effect on multiple outcomes. Once an appropriate $\bar{\theta}_a$
value for the target magnitude of treatment effect is determined, one is
ready to set up hypotheses, compute the sample size needed, and test the
treatment global effect using O'Brien's GST after the data are
collected. Determination of global treatment effect, computation of
sample size and testing of GTE using functions included in this package
are illustrated in Part I, II and III respectively.

## Data preparation

Data has to been consistently coded to ensure that larger values
indicate better outcomes for all measures.

![Data Structure: The K Outcomes are Denoted by V1, V2, ..., and VK. The
Entries are Observations from Patients](images/data.png)

## Part I. Determination of the global treatment effect

This section describes how to compute the global treatment effect
$\bar{\theta}$ from pilot data (if they are available) or summary
statistics (if pilot data is not available).

### Global treatment effect estimation when there are pilot data

When there are pilot data, we can directly use the theta function in
this package to get an estimation of global treatment effect.

Suppose $X1$ holds observations from group 1 and $X2$ holds observations
from group 2, in which data are consistently coded. Then you can
directly pass $X1$ and $X2$ into theta function in form of theta(X1,
X2). This function will then calculates the estimated treatment effect
for this outcome.

Example:

x1 is 100 random samples from a t-distribution with 50 degrees of
freedom, while x2 is 100 random samples from a t-distribution with 2
degrees of freedom. We treat x1 as outcomes from group 1 and x2 as
outcomes from group 2 in pilot data. The $\bar{\theta}$ is calculated as
follow:

```{r}
set.seed(123)
x1<- rt(100, 50, ncp = 10)
x2<- rt(100, 2, ncp = 10.3)
# use original data:
theta_1 <- theta(x1,x2) 
theta_1
# effect size when comparing x1 and x2:
effect_size1 <- abs(mean(x1)-mean(x2))/sqrt(var(c(x1,x2)))
```

Under this scenario, the global treatment effect $\bar{\theta}$ =
0.3056, indicating that group 2 gives globally higher values than group
1.

### Global treatment effect estimation when there is no pilot data

When there are no pilot data, investigators can estimate the treatment
effect based on information from the literature. In most cases, such
information is presented in terms of mean and standard deviation using
normal approximation, or presented using summary statistics of other
parametric distributions.

In some studies, nonlinear monotone transformations (eg., log
transformation, square root transformations, etc.) were applied to the
data to obtain a better parametric approximation, and only summary
statistics of the transformed data were reported. Fortunately, this is
not a problem when GTE is used to measure a treatment's effect since the
GTE is unchanged as it is computed only from the observation ranks.

Example:

Suppose we have hypothetical summary statistics using normal
approximation as shown below.

```{r echo=FALSE}
mu1 <- c(190,80,50) # mean from treatment 1
stdev1 <- c(90,30,30) # standard deviation from treatment 1
mu2 <- c(200,70,30) # mean from treatment 2
stdev2 <- c(80,30,30)

summary_data <- data.frame(
  Outcome_Scale = c("Outcome 1", "Outcome 2", "Outcome 3"),
  Placebo_Mean = mu1,
  Placebo_SD = stdev1,
  Treatment_Mean = mu2,
  Treatment_SD = stdev2
)
ex_summary <- kable(summary_data, 
      col.names = c("Outcome scale", "Placebo Mean", "Placebo SD", "Treatment Mean", "Treatment SD"), 
      align = 'c')
ex_summary
```

We can calculate the treatment effect $\theta$ of individual outcomes
based on
$\theta = 2\Phi\left(\frac{\mu_1 - \mu_2}{\sqrt{\sigma_1^2 + \sigma_2^2}}\right) - 1$
[@GTE]

```{r}
theta <- 2*pnorm((mu2-mu1)/sqrt(stdev1^2+stdev2^2))-1
theta
```

For outcome 1, group 2(treatment) gives globally smaller values than
group 1 (control). For outcome 2, there is no global difference between
group 1 and 2. For outcome 3, group 2 gives globally higher values than
group 1.

\*\* summary statistics in form of other distribution

## Part II. Determination of sample size

### Sample size estimation when some historical data are available

We could use N.GSTwPilotData function to derive the sample size when
there is some pilot data.

Functions required inputs are:

1\. data: a (m1+m2) x k matrix of pilot data with its first m1 rows from
treatment 1, and the rest m2 rows from treatment 2. k is the number of
endpoints

2\. m1: number of patients in treatment 1 in the pilot data

3\. r0: randomization ratio in new study = n2/n1

4\. test.side: 1 (one-sided test) or 2 (two-sided test)

5\. alpha: type I error to be controlled

6\. power: statistical power to be controlled

7\. GTE: global treatment effect where power is controlled at.


Example: Suppose there are 3 endpoints in this hypothetical historical dataset. 
The first 40 rows are random observations from treatment 1 which follow standard
normal distribution. The rest 60 rows are random observations from
treatment 2 which follow normal distribution with mean 0.6.


```{r echo=FALSE}
# hypothetical pilot data
set.seed(123)
k<- 3
n1<- 40
N<- 100
n2<- N-n1
x1<- rnorm(N)
data<- cbind(x1/sqrt(2)+rnorm(N)/sqrt(2),
              x1/sqrt(2)+rnorm(N)/sqrt(2),
              x1/sqrt(2)+rnorm(N)/sqrt(2))
x2<- rnorm(n2, mean=0.6)
data2<- cbind(x2/sqrt(2)+rnorm(n2)/sqrt(2),
              x2/sqrt(2)+rnorm(n2)/sqrt(2),
              x2/sqrt(2)+rnorm(n2)/sqrt(2))
data[(n1+1):N,]<- data[(n1+1):N,]+data2
head(data) 
```

Now, suppose the goal is to attain 0.05 significance level with 90%
power in two sided hypothesis testing. We pass these requirements into
the function:

```{r}
# calculate sample size
N.GSTwPilotData(data, m1=n1, 
  r0=n2/n1, test.side=2, alpha=0.05, power=0.9, GTE=NA)
```

As the result shows, we would need 66 participants in group 1 and 100
participants in group 2.

### Sample size estimation when no historical data are available

We could use N.GSTnoData function to derive the sample size when there
is no pilot data.

Functions required inputs are:

1\. k: number of endpoints

2\. r0: randomization ratio = n2/n1

3\. GTE: target GTE where power is controlled

4\. sigma2: upper bound of variances F1(X2) and F2(X1)

5\. rho: Spearman rank correlation

6\. alpha: type I error

7\. power: desired statistical power

8\. test.side: 1(one-sided test) or 2 (two-sided test)

Using the same hypothetical summary statistics using normal
approximation before:

```{r echo=FALSE}
ex_summary
```

In previous section, we obtained the treatment effect $\theta$ for each
outcome, we can then calculate the global treatment effect by taking the
average $\bar{\theta}$ of these $\theta$ values.

We can also derive an estimation of the standard deviation
$\sigma^2 = \min\left\{\frac{1}{12}, \frac{1 - \min_{1 \leq v \leq K} (\theta_va^2)}{4}\right\}$.
[@samplesize]

Plug these values into the function, we can get sample size estimation
for each group.

```{r}
GTE<- mean(theta)
sigma2<- min((1-min(theta^2))/4, 1/12) 
# rest of the variables are pre-determined:
k<- length(mu1) # number of endpoints
alpha<- 0.05 # type I error
beta<- 0.1 # 90% power
test.side<- 1
r0<- 1 # randomization ratio
rho<- 0.1
N.GSTnoData(k, r0, GTE, sigma2, rho, alpha, beta, test.side)
```

### Compare to univariate analysis

One advantages of using GTE is that it helps investigators to obtain the
desired power and significance level with smaller sample size compared
to univariate analysis.

Here, we estimate the sample sizes needed in each arm of the group based
on univariate analysis to obtain 90% power and 0.05 significance level
using the same hypothetical summary data.

```{r}
N.Bon <- rep(NA, k)
for (i in 1:k){
  d1<- abs(mu1[i]-mu2[i])/ max(stdev1[i],stdev2[i])
  x1<- pwr.t.test(n = NULL, d = d1, power=0.9, sig.level=0.05/3,
                  type="two.sample", alternative="greater")
  N.Bon[i]<- round(x1$n)
}
N.Bon
```

We would need 1884 participants in each arm in outcome 1, 210
participants in each arm in outcome 2, and 53 participants in each arm
in outcome 3.

As the result shows, sample sizes based on univariate analysis are much
greater compared to sample size estimation based on GTE.

## Part III. Use of global statistical testing in data analysis

In previous sections, we've determined appropriate $\bar{\theta_a}$ for
the target magnitude of treatment effect, set up the hypotheses,
computed the sample size needed, and now, we can test the treatment's
global effect by calculating the p-value using GST() function.

```{r}
GST(data=data, m=n1, test.side = 1)
```

Combine the fact that the GTE value is 0.231, which is greater than 0 and the fact that p-value is 0.005, which is smaller than significance level of 0.05, we conclude that there is significant treatment benefit.

GST.simu

```{r}
m<- 50 # sample size in treatment group 1
n<-  50 # sample size in treatment group 2
k<- 20 # number of endpoints
r<- rep(1, k)
rho<- 0.01 # correlation
simu<- 5

i<-1
dataset<- array(rnorm((m+n)*k*simu,0,1) , dim=c(m+n,k, simu))
 for (j in 2:k) #introduce correlation for group 1
 {dataset[1:m,j,]<- sqrt(rho)*dataset[1:m,1,]+sqrt(1-rho)*dataset[1:m,j,]
 }
dataset[(m+1):(m+n),1,]<- matrix(rnorm(n*simu,0,r[1]), nrow=n,ncol=simu)
for (j in 2:k) #introduce correlation for group 2
{ dataset[(m+1):(m+n),j,]<- sqrt(rho)*dataset[(m+1):(m+n),1,]+sqrt(1-rho)*
                       matrix(rnorm(n*simu,0,r[j]), nrow=n,ncol=simu)
}
delta<- 0.15 # introduce difference between group 1 and group 2
dataset[(m+1):(m+n),,]<- dataset[(m+1):(m+n),,]+delta
dim(dataset) #[1] 100  20   5

# GST from simulated data:
x<- GST.simu(data=dataset,m)
```

## References

[@Huang2005]
